Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-23 paper.tex
Added a subsection outlining the document

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the
drugs we take are the result of intensive and costly research, and the
more information is shared (a process sometimes very hard to achieve
in pharmaceutical companies) the easier it should be to develop
novel treatments. How chemical information relates to computer
science becomes clear when realizing the amount of data available
currently---74 million different molecules are currently stored in
public databases (such as PubChem) with a large amount of annotations,
and the design of data structures for chemicals, as well as subsequent
data mining of this information, are areas where expertise in computer
science is dearly needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

@@

Graph mining algorithms can be applied to this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (although not in others).  Nevertheless, the
3D information is lost when one only considers the graph of a
molecule. To extend the concept of three-dimensionality further, in
many biological systems the \emph{surface properties} of the molecule
are what appear to be relevant, as shown in the bottom right hand
corner.  Hence, while graph mining on chemicals is certainly a
promising avenue to explore, one needs to keep in mind that not all
molecular properties are encoded in the graph (for example, neither
rotational states, nor different charge states are encoded) and thus
often other `descriptors' need to be used to capture the essentials of
a chemical system \cite{bender2004}.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity to collect, store, and mine data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data by bridging the gap between risk in
generalizable encodings and complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

\subsection{Databases \& Ontologies}

@@

\subsection{Activity Mining \& Prediction}
\label{sec:activity-mining-}

@@

\subsection{Open Source Chemical Toolkits}

>>>>

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discuss later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, and the more
information is shared (a process sometimes very hard to achieve in
pharmaceutical companies) the easier it should be to develop novel
treatments. How chemical information relates to computer science
becomes clear when realizing the amount of data available
currently---74 million different molecules are currently stored in
various public databases (such as PubChem) with a large amount of
annotations, and the design of data structures for the representation
and manipulation of chemical structure, as well as subsequent data
mining of this information, are areas where expertise in computer
science is dearly needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

@@

Graph mining algorithms can be applied to this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (although not in others).  Nevertheless,
the 3D information is lost when one only considers the graph of a
molecule. To extend the concept of three-dimensionality further, in
many biological systems the \emph{surface properties} of the molecule
are what appear to be relevant, as shown in the bottom right hand
corner.  Hence, while graph mining on chemicals is certainly a
promising avenue to explore, one needs to keep in mind that not all
molecular properties are encoded in the graph and thus often other
`descriptors' need to be used to capture the essentials of a chemical
system \cite{bender2004}.

From the above it becomes clear that the nature of many problems in
cheminformatics require that a complex mixture of combinatorial and
continuous techniques.  If we oversimplify the nature of chemical
structures and their properties, we can not describe the world around
us in sufficient detail. On the other hand, if we try to encode every
detail (\emph{e.g.}, all possible 3D structures and their properties,
all metabolites of a structure in humans in different tissues), the
complexity of collecting, storing, and mining data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data by bridging the gap between risk in
generalizable encodings and complexity (space and time). The goals of
cheminformatics are to support better chemical decision making by
1. storing and integrating data in maintainable ways, 2.  providing
enough open standards and tools so software engineering concepts allow
applications and data to be used across heterogeneous platforms, and
3. mining the many chemical property spaces in a time- and
space-efficient way. We will discuss all three goals in the following
sections.

\subsection{Outline}
\label{sec:outline}

The remainder of the article discusses various topics in
cheminformatics by placing them in the context of the relevant
subfields of computer science, aiming to highlight the challenges that
need to be addressed and possible entry points for computer science
research. Section \ref{sec:databases} discusses issues related to the
storage of chemical structures in relational databases and how
ontologies allow us to query such collections beyond the standard
graph based subgraph matching approaches. While databases usually
contain explicit molecules, enumeration of sctructures based on
various constraints is key to a variety of chemical problems ranging
from mass spectrometry to reaction planning. Section
\ref{sec:struct-enum} discusses challenges in this area. Section
\ref{sec:activity-mining} highlights the problem of making predictions
of biological activity from chemical structure and tieing it all
together, Section \ref{sec:knowledge-management} discusses the
challenges of managing chemical structure data along with associated
biological data in production scenarios. Finally Section
\ref{sec:role-open-source} discusses the public sources of chemical
information and tools available for the handling of chemical structre
-- both of which are required for for the development of
cheminformatics research programs.

@@

\subsection{Databases \& Ontologies}
\label{sec:databases}

@@

\subsection{Activity Mining \& Prediction}
\label{sec:activity-mining}

@@

\subsection{Open Source Chemical Toolkits}
\label{sec:toolkits}
